Evolution of ORM Systems: Formal Foundations

### 2012-12-22 doc/description.tex
orm algorithm

\documentclass[11pt]{article}
\usepackage{amssymb}
\usepackage{amsmath}
\usepackage{semantic}
\usepackage{listings}
\usepackage{hyperref}
\usepackage{graphicx}

@@

\section{Data Evolution in ORM Software}
\label{sec:evolution}
The ORM software consist of an application created using object oriented language and relational database. The application and database are in consistent state according to a concrete ORM function, otherwise the software cannot work properly. An ORM software is illustrated in Figure \ref{fig:appStructure}. The software can be defined as a triple:
$$software = < Application, Database, \rho >$$
Models of $Application$ and $Database$ will be introduced in detail in Section \ref{sec:appModel} and in Section \ref{sec:dbModel}, the ORM function $\rho$ is defined in Section \ref{sec:orm}. The consistent state of application and database is defined as:
$$\rho(Application) = Database$$

@@

A row in a table is tuple consisting of reference to a concrete table and a set of value pairs, which represents concrete values of concrete columns.
$$row = < refTable, pair* >$$
$$pair = < reference, value >$$
The $reference $ is a name of a column, a foreign key or its primary key and $value$ represents concrete value stored in the database. Each row in a database has to refer to a table in database schema, otherwise the database is inconsistent. The consistency of database is then:
\begin{equation*}
\begin{gathered}
\forall r \in data(d), \exists t \in tables(d) : reference(r) = name(t) \\ \wedge \forall p \in pairs(r) \exists u \in columns(t) \cup foreignKeys(t) : reference(p) = name(u) 
\end{gathered}
\end{equation*}

\begin{itemize}
	\item \textbf{TODO: define operations such as instantiated}
\end{itemize}

$instantiated : Class \times Application \rightarrow Boolean $ \\

\section{Object-Relational Mapping}
\label{sec:orm}
There are defined functions for object-relational mapping used in the model in this section. The purpose of a function is indicated in its name and should be obvious from its definition. These functions do not serve for full object-relational mapping, their purpose is to help with data evolution (see Section \ref{sec:eorm}) and its propagation from application to database.

\begin{itemize}
	\item \textbf{TODO: $\rho(Application) = Database$ show how database is build from Application}
\end{itemize}

@@

\begin{center}
$ \rho_{t} : AppType \rightarrow DbType $ \\
$ \rho_{t}(APPSTRING) = DBSTRING $
\end{center}

@@

\subsubsection{Application Alternation}
\paragraph{Rename}
\paragraph{Change Cardinality}
\paragraph{Copy Property}

@@

\subsubsection{Creation of a System}
\paragraph{Add Class} The $AddClass$ operation changes the structure of the database schema only.
$$
\Phi_{AC}(c, d) = \inference{ \forall t \in tables(d) : name(t) \neq name(\rho(c)) \\
	\forall a \in associations(c), \exists u \in  tables(d) : reference(a) = name(u) 
}{
\begin{gathered}
  \rho(c) \wedge \rho(singleProperties(c)) \wedge \rho(collectionProperties(c)) \\ \wedge \rho(associations(c))
\end{gathered}
}
$$

@@

\subsubsection{Deconstruction of a System}
\paragraph{Remove Class}
$$
\Phi_{RC}(c,d) = \inference{ \rho(c) \in tables(d) \\ \wedge \not \exists r \in data(d) : owningTable(r) = \rho(c) \\ \not \exists j \in tables(d), f \in foreignKeys(j) : reference(f) = \rho(c)  }{ tables(d) \setminus \rho(c)}
$$
%TODO remove by label

@@

\subsubsection{Altering a System}

>>>>

\documentclass[11pt]{article}
\usepackage{amssymb}
\usepackage{amsmath}
\usepackage{semantic}
\usepackage{listings}
\usepackage{hyperref}
\usepackage{graphicx}
\usepackage{algpseudocode}

@@

\section{Data Evolution in ORM Software}
\label{sec:evolution}
The ORM software consist of an application created using object oriented language and relational database. The application and database are in consistent state according to a concrete ORM function, otherwise the software cannot work properly. An ORM software is illustrated in Figure \ref{fig:appStructure}. The software can be defined as a triple:
\begin{equation*}
	software = < Application, Database, \rho >
\end{equation*}
Models of $Application$ and $Database$ will be introduced in detail in Section \ref{sec:appModel} and in Section \ref{sec:dbModel}, the ORM function $\rho$ is defined in Section \ref{sec:orm}. The consistent state of application and database is defined as:
$$\rho(Application) = Database$$

@@

A row in a table is tuple consisting of reference to a concrete table and a set of value pairs, which represents concrete values of concrete columns.
$$row = < refTable, pair* >$$
$$pair = < reference, value >$$
The $reference $ is a name of a column, a foreign key or its primary key and $value$ represents concrete value stored in the database. Each row in a database has to refer to a table in database schema, otherwise the database is inconsistent. The consistency of database is then:
\begin{equation*}
\begin{gathered}
\forall r \in data(d), \exists t \in tables(d) : reference(r) = name(t) \\ \wedge \forall p \in pairs(r) \exists u \in columns(t) \cup foreignKeys(t) : reference(p) = name(u) 
\end{gathered}
\end{equation*}
There are two functions connected with stored data - first of them verify if the table contains any data:
$$contains : Table \times Database \rightarrow Boolean $$
\begin{equation*}
\begin{gathered}
	contains(t, d) = \begin{cases}
 \inference{\exists r \in data(d) : refTable(r) = name(t) }{true} \\ \\
  \inference{\forall r \in data(d) : refTable(r) \neq name(t)}{false}
 \end{cases} \\ \\
 t \in tables(d), d \in Database
\end{gathered}
\end{equation*}

second of them verify if there are instances of a class:
$$instantiated : Class \times Software \rightarrow Boolean $$ 
\begin{equation*}
\begin{gathered}
	instantiated(c, s) = \begin{cases}
 \inference{\exists t \in tables(database(s)) : name(t) = name(c) \\ \wedge contains(t, database(s))}{true} \\ \\
  \inference{\exists t \in tables(database(s)) : name(t) = name(c) \\ \wedge ! contains(t, database(s))}{false} \\ \\
 \end{cases} \\ \\
 c \in classes(application(s)), s \in Software 
\end{gathered}
\end{equation*}

\section{Object-Relational Mapping}
\label{sec:orm}
There are defined functions for object-relational mapping used in the model in this section. The purpose of a function is indicated by its name and should be obvious from its definition. These functions do not serve for full object-relational mapping, their purpose is to help with data evolution (see Section \ref{sec:eorm}) and its propagation from application to database. However the database can be obtained from an application model by using following algorithm:

\begin{algorithmic}[1]
	\Require $a \in Application$
	\State d = Database
	\ForAll{$c \in classes(a)$} 
		\State $\rho_c(c), d$
		\ForAll{$p \in properties(c)$} 
			\State $\rho_p(p, d)$
		\EndFor
	 \EndFor
	\ForAll{$c \in classes(a)$}
		\ForAll{$as \in associations(c)$} 
			\State $\rho_{a}(as, d)$
		\EndFor
	\EndFor
\end{algorithmic}

@@

\begin{equation*}
\begin{gathered}
	\rho_{t} : AppType \rightarrow DbType  \\
 	\rho_{t}(APPSTRING) = DBSTRING 
\end{gathered}
\end{equation*}

@@

\subsubsection{Application Alternation}
\paragraph{Rename Class}
$$\rho_{RenC} : Label \times Label \times Application \rightarrow Class $$
\begin{equation*}
	\rho_{RenC}(l_o, l_n, a) = \inference{\exists c \in classes(a) : name(c) = l_c \\ \wedge \forall e \in classes(a) : name(e) \neq l_n}{\begin{gathered}
c.name = l_n \wedge \forall f \in associations(e), \\ e \in classes(a), reference(f) = l_c : f.reference = l_n 
\end{gathered}
}
\end{equation*}

\paragraph{Rename Property}
$$\rho_{RenP} : Property \times Label \times Application \rightarrow Property $$
\paragraph{Rename Association}
$$\rho_{RenA} : Association \times Label \times Application \rightarrow Association $$
\paragraph{Change Cardinality}
$$\rho_{CcP} : Property \times NzNat \times Application \rightarrow Property $$
$$\rho_{cardA} : Association \times NzNat \times Application \rightarrow Association $$
\paragraph{Copy Property}
$$\rho_{copyP} : Property \times Class \times Class \times Application \rightarrow Class \times Class $$

@@

\subsubsection{Creation of a Software}
\paragraph{Add Class} The $AddClass$ operation changes the structure of the database schema only.
$$
\Phi_{AC}(c, d) = \inference{ \forall t \in tables(d) : name(t) \neq name(\rho(c)) \\
	\forall a \in associations(c), \exists u \in  tables(d) : reference(a) = name(u) 
}{
\begin{gathered}
  \rho(c) \wedge \rho(singleProperties(c)) \wedge \rho(collectionProperties(c)) \\ \wedge \rho(associations(c))
\end{gathered}
}
$$

@@

\subsubsection{Deconstruction of a Software}
\paragraph{Remove Class}
$$
\Phi_{RC}(c,d) = \inference{ \rho(c) \in tables(d) \\ \wedge \not \exists r \in data(d) : owningTable(r) = \rho(c) \\ \not \exists j \in tables(d), f \in foreignKeys(j) : reference(f) = \rho(c)  }{ tables(d) \setminus \rho(c)}
$$
%TODO remove by label

@@

\subsubsection{Altering a Software}
